\documentclass[phdthesis,12pt]{wuthesis}

\usepackage{url}
\usepackage{hyperref}
\usepackage{tikz}
\usepackage{float}
\usepackage{graphicx}
\usepackage{cprotect}
\usepackage{amsmath}
\usepackage{longtable}
\usepackage{fancyvrb}
\usepackage{subscript}
\usepackage{hyperref}
\usepackage{fixltx2e}

\usetikzlibrary{shapes, calc, patterns, arrows, decorations.pathreplacing, decorations.markings, positioning, automata}

\newcommand{\rcgo}{rc\textsubscript{go}}

\renewcommand{\thesisauthor}{Justin Wilson}
\renewcommand{\thesisauthorlastname}{Wilson}

\renewcommand{\thesisdepartment}{Department of Computer Science and Engineering}
\renewcommand{\thesisfield}{Computer Science}

\renewcommand{\thesismonth}{December}
\renewcommand{\thesisyear}{2016}

\renewcommand{\thesistitle}{A Transactional Model and Platform for Designing and Implementing Reactive Systems}
\renewcommand{\thesisshorttitle}{Reactive Components}

\renewcommand{\thesissupervisor}{Professor Christopher Gill}
\renewcommand{\thesiscommittee}{Christopher Gill, Chair \\
  Kunal Agrawal \\
  Roger Chamberlain \\
  Ron Cytron \\
  Dave Peters \\
  Gruia-Catalin Roman}

%% Enter the dedication
%% \renewcommand{\thesisdedication}{Dedicated to my parents.}

\begin{document}
\VerbatimFootnotes

\frontmatter
\begin{thesistitlepage}
%\iffalse
\end{thesistitlepage}

\begin{thesiscopyrightpage}
%\iffalse
\end{thesiscopyrightpage}

\begin{singlespace}
\setcounter{page}{2}
\tableofcontents
\listoftables
\listoffigures
\end{singlespace}

\begin{thesisacknowledgments}

\begin{quote}
Blessed be the God and Father of our Lord Jesus Christ, the Father of mercies and God of all comfort  \emph{2 Corinthians 1:3}
\end{quote}

First and foremost, I would like to thank my wife who has been beside me every step of the way.

I am forever grateful for the patience, guidance, and wise counsel of my advisor Dr. Gill.

The seed for this dissertation was planted by Dr. Roman in the basement of Lopata Hall during a course on UNITY.
He was lecturing on the development of structured programming and the give and take between formal methods and practice.
I am grateful to Dr. Roman for introducing me to the clarity of thinking that comes from formal models.

I would like to thank the Department of Computer Science and Engineering, the National Science Foundation, and the Ford Motor Company for their generous financial support.

\end{thesisacknowledgments}

%% \begin{thesisdedicationpage}                %% Generate the dedication page.
%% %\iffalse
%% \textbf{Note:} You may include a special dedication as shown here.  If you
%% include this page, be sure to keep it brief and center it on the page both
%% horizontally and vertically.  Alternatively, you may remove this page
%% altogether, and a special dedication can be placed as the final paragraph to
%% your acknowledgments page (as shown in this document on the preceding page).
%% %\fi
%% \end{thesisdedicationpage}

\begin{thesisabstract}
\input abstract.tex
\end{thesisabstract}

\mainmatter
\include{introduction}
\include{background}
\include{model}
\include{language}
\include{implementation}
\include{scheduler}
\include{conclusion}

\appendix

\chapter{Partition Tables}
\label{partition_tables}

\begin{longtable}{ccccccccr}
Symbol & System & Server & Counter & Client & Response & Tick & Request & Count \\
\hline
\endhead
\input{async_partitions.tex}
\caption[Partitions for the AsyncClock system]{Partitions for the AsyncClock system.  The Symbol column contains the symbol used to represent this partition on plots.  The System, Server, Counter, and Client columns indicate the thread used for the garbage collection action for the respective component.  The Response, Tick, and Request columns indicate the thread used for the respective transaction.  The Count column indicates the number of samples for this partition.}
\label{async_partitions}
\end{longtable}

\begin{longtable}{ccccccccr}
Symbol & System & Counter & Request & Tick & Count \\
\hline
\endhead
\input{sync_partitions.tex}
\caption[Partitions for the SyncClock system]{Partitions for the SyncClock system.  The Symbol column contains the symbol used to represent this partition on plots.  The System and Counter columns indicate the thread used for the garbage collection action for the respective component.  The Request and Tick columns indicate the thread used for the respective transaction.  The Count column indicates the number of samples for this partition.}
\label{sync_partitions}
\end{longtable}

\bibliographystyle{plain}
\begin{spacing}{1.0}
\bibliography{paper}{}
\end{spacing}
\nocite{*}

\begin{thesisauthorvita}
%
% This is just a sample of what to do in a vita
%
\begin{center}
{\large\thesisauthor}
\end{center}
%
% Need to 'fold' long labels
%
\newcommand{\vitalabel}[1]%
  {\raisebox{0pt}[1ex][0pt]
    {\makebox[\labelwidth][l]%
      {\parbox[t]{\labelwidth}{\hspace{0pt}\textbf{#1}}}}}
%
% Here's the list definition
%
\begin{list}
  {}%                                        nodefault label
  { \renewcommand{\makelabel}{\vitalabel}%   setting labels
    \setlength{\labelwidth}{100pt}%          label width
    \setlength{\leftmargin}{120pt}%
    \setlength{\itemindent}{0pt}%            don't indent first line of items
    \setlength{\parsep}{\baselineskip}%               space between paragraphs
    \setlength{\itemsep}{5pt}%               additional space between items
    }

\item[Degrees] B.S.\ Summa Cum Laude, Electrical Engineering, December 2006 \\
               B.S.\ Summa Cum Laude, Computer Engineering, December 2006 \\
               Ph.D. Computer Science, December 2016
%% \item[Professional\linebreak Societies]
%%   Association for Computing Machines \\
%%   The Touring Society \\
%%   The Free Software Foundation
\item[Publications]
  Wilson, J., Dai, M., Jakupovic, E., Watson, S., \& Meng, F. (2007). Supercomputing with toys: harnessing the power of NVIDIA 8800GTX and playstation 3 for bioinformatics problems. In \textit{Computational Systems Bioinformatics Conference} (Vol. 6, pp. 387-390).

  Thomas, L., Wilson, J., Roman, G. C., \& Gill, C. (2009, November). Achieving coordination through dynamic construction of open workflows. In \textit{ACM/IFIP/USENIX International Conference on Distributed Systems Platforms and Open Distributed Processing} (pp. 268-287). Springer Berlin Heidelberg.

  Xi, S., Wilson, J., Lu, C., \& Gill, C. (2011, October). RT-Xen: towards real-time hypervisor scheduling in xen. In \textit{IEEE International Conference on Embedded Software (EMSOFT)} (pp. 39-48).
\end{list}
\flushright
\thesismonth\ \thesisyear
\end{thesisauthorvita}

%% \begin{thesisshorttitlepage}
%% %\iffalse
%% {\small

%% \textbf{NOTE:}	This is a sample of a ``short title'' page.  Please change the
%% line above to use an appropriate ``short title'' for your thesis, insert your
%% last name, and include your degree and year in which the degree will be earned.
%% Separate elements using commas, as illustrated in the sample above.
%% \uline{Your ``short title'' cannot exceed 35 characters, counting spaces}.  It
%% does not matter if there is a page number at the bottom of the page.

%% \textbf{IMPORTANT:} This page should be printed and taped securely to each of
%% the three manila envelopes used to submit your final hard copies.  Remove this
%% page before submitting your final copies (i.e., this page should not be
%% included in either your electronic submission or your hard copy submissions).
%% See Appendix~\ref{app:procedures} for further details if necessary.
%% }
%% %\fi
%% \end{thesisshorttitlepage}

\end{document}
